Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-07-09 papers/wdmerger_I/paper.tex
Add a description of our gravitational wave signal diagnostics to Paper I.

@@

The authors thank Noel Scudder and Platon Karpov for their help with 
this project, especially related to visualization of the results. 
We also thank Adam Jacobs for his advice on and assistance with 
running on supercomputing resources.

>>>>

A final diagnostic quantity we consider is the gravitational wave emission by the 
binary system. White dwarfs are not strongly affected by general relativistic effects;
the orbital motions are much slower than the speed of light, and the relativity parameter 
$GM / c^2 R$, which measures the ratio of the Schwarzschild radius of a mass $M$ to the actual 
radius $R$ of the object, is much less than unity for a white dwarf. It is justified to 
neglect any relativistic effects in computing the dynamical evolution of the system. However,
the system emits gravitational waves during its evolution; this energy loss is what drives 
the initial inspiral over very long timescales, and occurs at a greater rate during 
the coalescence phase, though then being inconsequential on a dynamical timescale. The 
frequency of the gravitational waves will be similar in size to the frequency of the 
orbital motion, which is in the range 10-100 mHz for our problem. This is well outside 
the range of currently existing gravitational wave detectors but is very well suited 
for proposed space-based detectors such as eLISA.

We follow the prescription of \citet{blanchet:1990}. At distances far from the 
gravitational wave source, we can consider the leading term in the gravitational 
wave signal:
\begin{equation}
  h^{TT}_{ij}(t,\mathbf{x}) = \frac{2G}{c^4 r}P_{ijkl}(\mathbf{n}) \ddot{Q}_{kl}(t - r/c).
\end{equation}
$h$ is the perturbation to the spacetime metric and is commonly called the \textit{strain}; 
for laser interferometers, it measures the relative change in the distance between mirrors. 
The ``TT'' superscript indicates that we work in the commonly used tranverse-traceless gauge.
This strain is measured at time $t$ and position $\mathbf{x}$ relative to the binary system.
$r\equiv |\mathbf{x}|$ is the distance from the observer to the binary system. The unit vector 
$\mathbf{n} \equiv \mathbf{x} / r$ then measures the direction of the outgoing wave with 
respect to the observer, and $P_{ikjl}(\mathbf{n})$ is an operator in the transverse-traceless 
gauge that projects a tensor onto the direction orthogonal to $\mathbf{n}$. $Q_{kl}$ is the 
quadrupole moment tensor:
\begin{equation}
  Q_{kl} = \int dV \rho \left(x^k x^l - \frac{1}{3}\delta^{kl} \mathbf{x}^2\right).
\end{equation}
The argument $(t - r/c)$ indicates that to get the strain at time $t$ we evaluate the second derivative of the 
quadrupole moment at the retarded time $t - r/c$. In practice we ignore this in calculating this strain,
instead using the current simulation time, and so this merely yields a time delay of order $r/c$ between 
when the waves are emitted and when we observe them.

Therefore the primary component of the calculation is the evaluation of the second time derivative of $Q_{kl}$.
Explicitly constructing a discretized form of this derivative, using the current state and the state at 
previous times, is undesirable because of the inherent imprecision (its accuracy depends on the size of the timestep),
in addition to the logistical challenges that may be implied by saving and using previous simulation states. 
\citet{blanchet:1990} provide a prescription for this time derivative purely in terms of the state at a given time:
\begin{equation}
  \ddot{Q}_{kl} = \text{STF}\left\{2\int dV \rho (v^k v^l + x^k g^l)\right\}.
\end{equation}
The symmetric trace-free (STF) operator is defined as:
\begin{equation}
  \text{STF}\left\{A^{ij}\right\} = \frac{1}{2}A^{ij} + \frac{1}{2}A^{ji} - \frac{1}{3} \delta^{ij} \sum_{k}A^{kk}.
\end{equation}

The strategy is then as follows. At the end of the coarse timestep, we first calculate $\ddot{Q}_{kl}$
using an integral over the domain (using refined grids as usual by masking out the contributions from 
coarse grids underlying more accurate fine grids). This quantity is independent of the observer. Then, 
we pick an observing location $\mathbf{x}$ relative to the domain, evaluate the projection operator, 
and then perform the relevant tensor contraction to determine the strain tensor. We can 
repeat this process for any number of observing locations at minimal cost, since the quadruple tensor 
only needs to be calculated once. Now, gravitational waves only excite modes orthogonal to their 
direction of travel. These are the ``plus'' and ``cross'' modes, $h_+$ and $h_\times$, named after 
the types of spatial distortions they exhibit. We calculate the signal at a distance $r$ along 
the $x$, $y$ and $z$ axes. For the latter, as an example, $h_{+} = h_{11} = -h_{22}$ and 
$h_{\times} = h_{12} = h_{21}$. All other entries vanish. By default we take $r = 10$ kpc; 
as shown by \citet{loren-aguilar:2005}, this is a typical distance scale over which an 
experiment such as LISA could detect a coalescing binary white dwarf system. 
However, the strain at any other distance is easily calculated and goes as the inverse of the distance.

@@

The authors thank Noel Scudder and Platon Karpov for their help with 
this project, especially related to visualization of the results. 
We also thank Adam Jacobs for his advice on and assistance with 
running on supercomputing resources. Special thanks are given 
to the organizers of the 2015 Caltech Gravitational Wave 
Astrophysics School, and their supporters: the NSF 
under CAREER award PHY-1151197, the Sherman Fairchild 
Foundation, the LIGO Laboratory at Caltech, and Caltech.
